---
title: "CLass 9: Machine Learning"
author: "Linh-An Ho"
date: "2/4/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

The main k-means function in R is called `kmeans()`. Let's play with it here.

```{r}
 # Generate some example data for clustering (30 points, centered round -3/3)
tmp <- c(rnorm(30,-3), rnorm(30,3)) 

# Taking the vector of 60 things and combining the data from tmp & rev(tmp) (reverses the order of tmp)
x <- cbind(x=tmp, y=rev(tmp))

plot(x)

```


Use the kmeans() function setting k (centers) to 2 and nstart=20 

```{r}
km <- kmeans(x, centers = 2, nstart = 20)
```

Inspect/print the results
```{r}
km
```


> Q. How many points are in each cluster?

30 

> Q. What ‘component’ of your result object details

- cluster size? size
- cluster assignment/membership? cluster

```{r}
# How many points are there in all
length(km$cluster)

# Tells you how many of each (i.e. 1s and 2s) there are 
table(km$cluster)
```

- cluster center? cluster means

Plot x colored by the kmeans cluster assignment and
      add cluster centers as blue points
```{r}
#  Cluster assignment numbers (1 and 2) align with the color number (1=black, 2=red)
plot(x, col = km$cluster)

# Can change from the default by adding +(number)
plot(x, col = km$cluster+1)

# Adding the cluster centers in blue
plot(x, col = km$cluster)
points(km$centers, col = "blue", pch =15)
```


## Hierarchical Clustering in R

The main hierarchical clustering function in R is called `hclust()`
An important point here is that you have to calculate the distance matrix deom your input data before calling `hclust()`

For this we will use the `dist()` function first
```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

Folks often view the results of hierarchical clustering graphically. 
From this graph, we get two "branches" one for each cluster (1-30) and one for (31-60)
```{r}
plot(hc)
# Draw a line across
abline(h=6, col="red", lty=2)
abline(h=4, col="blue")
```

To get cluster membership vector I need to "cut" the tree at a certain height to yield my separate cluster branches
```{r}
# Cut by height h
cutree(hc, h=6)
```

```{r}
cutree(hc, h=4)
```

```{r}
# Can tell it to cut into desired amount of clusters
gp4 <- cutree(hc, k=6)
```



```{r}
 # Step 1. Generate some example data for clustering
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), 
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2),  
  matrix(c(rnorm(50, mean = 1, sd = 0.3) ,
  rnorm(50, mean = 0, sd = 0.3)), ncol = 2)) 
colnames(x) <- c("x", "y")

# Step 2. Plot the data without clustering
plot(x)

# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results) 

col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
d <- dist(x)
hc <- hclust(d)
plot(hc)
abline(h=2, col="red")
```

Use `cutree()` to get he clustership vector, and `table()` to see how many of each are in the cluster
```{r}
# Get cluster membership vector
grps <- cutree(hc, k=3)
table(grps)

# Plot the membership vector 
plot(x, col=grps)
```


##PCA of UK food data

```{r}
x <- read.csv("UK_foods.csv", row.names = 1)
x
```

Making plots to explore the data

Barplot
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

Pairs Plot: pairs one country on one axis and another on the other axis; good for small datasets. 
The farther the points from the middle line, the more different those data points are from each other
```{r}
pairs(x, col=rainbow(10), pch=16)
```

Principal Component Analysis (PCA): `prcomp()`function.
"t" rotates the data set to switch rows and columns
```{r}
pca <- prcomp( t(x))
#pca
```

`attributes()` tells us what are the components available in the results object (i.e. pca object)
```{r}
attributes(pca)
```

```{r}
# plotting PCA1 and PCA2
plot(pca$x[,1], pca$x[,2], xlab= "pca1", ylab= "pca2")
text(pca$x[,1], pca$x[,2], colnames(x), col=c("black", "red", "blue", "dark green"))
```


```{r}
summary(pca)
```
















