---
title: 'Class 6: R Functions'
author: "Linh-An Ho"
date: "1/23/2020"
output: github_document
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions 
read.table("Your file name", Header = TRUE, sep = "\t")

## Level 2 heading

### Level 3 heading

```{r}
# this is a silly plot!
plot(1:5)
```

Let's see more about **file import** (i.e. reading files into R). The main read function in base R is `read.table()`


```{r}
t1 <- read.table("test1.txt")
```

We need to add arguments to get this file imported

```{r}
t1 <- read.table("test1.txt", header = TRUE, sep = ",")
```

Or I could just use `read.csv()` which has the arguments I want in this case!
```{r}
t1 <- read.csv("test1.txt")
t1
```

Test 2
```{r}
t2 <- read.table("test2.txt", header = TRUE, sep = "$")
t2
```

Test 3
```{r}
t3 <- read.table("test3.txt")
t3
```

# Back to functions

Our first example function:

```{r}
add <- function(x, y=1) {
# Sum the input x and y x+y
  x + y
}
```

Let's try using ths function

```{r}
add(7, 3)
```

How does this work with vectors as input

```{r}
add( c(1, 2, 4))
```

```{r}
add( c(1, 2, 4), 4)
```

```{r}
add( c(1, 2, 4), c(1,2,4))
```

What is this `range()` function?
```{r}
x <- c(4,4,10,3,11)
max(x)
min(x)
```

```{r}
range(x)
```

This is our second function:
```{r}
rescale <- function(x) {
rng <-range(x)
(x - rng[1]) / (rng[2] - rng[1])
}
```

```{r}
rescale(x)
```

Do some more testing
```{r}
rescale(1:10)
```

```{r}
x <- c(1,2,NA,3,10)
rng <- range(x)
rng
```

Add na.rm to omit the "NA" form the dataset
```{r}
x <- c(1,2,NA,3,10)
rng <- range(x, na.rm = TRUE)
rng
```

```{r}
rescale2 <- function(x) {
rng <-range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
}
```

```{r}
rescale2(c(1,2,NA,3,10))
```

```{r}
rescale3 <- function(x, na.rm=TRUE, plot=FALSE) {
   rng <-range(x, na.rm=na.rm)
    print("Hello")

  answer <- (x - rng[1]) / (rng[2] - rng[1])
   return(answer)

  print("is it me you are looking for?")
  
  if(plot) {
plot(answer, typ="b", lwd=4)
}
print("I can see it in ...") 
return(answer)
}
```

Call the library once you've downloaded the package
```{r}
library(bio3d)
```

Hands on section
```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


> Q1. What type of object is returned from `read.pdb`

It is a list of 8 things and of class "pdb", "sse"
```{r}
class(s1)
```

```{r}
str(s1)
```

```{r}
aa321(s1$seqres)
```

```{r}
s1$atom
```

> Q2. What does the trim.pdb() function do?

It trims the information to a smaller subset

> Q3. What input parameter would turn off the marginal black and grey rectangles in the plots and what do they represent in this case?

"sse" would turn off the marginal black and grey rectangles; they represent whether the chain is a sheet, turn, or a helix.

> Q4. What would be a better plot to compare across the different proteins?

Combine plots into one
```{r}
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
points(s2.b, typ="l", col = "blue", lwd = 2)
points(s3.b,  typ="l", col = "red", lwd = 2)
```

> Q5. Which proteins are more similar to each other in their B-factor trends. How could you quantify this?

```{r}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```

```{r}
dist(rbind(s1.b, s2.b, s3.b))
```


> Q6. How would you generalize the original code above to work with any set of input protein structures?
  
  
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")  
  
  
```{r}
library(bio3d)

```

