---
title: 'Class 10: Unsupervised Learning Mini-Project'
author: "Linh-An Ho"
date: "2/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of Human Breast Cancer Cells

First we need to import (i.e. read) our input data.
```{r}
wisc.df <- read.csv("WisconsinCancer.csv")
head(wisc.df)
```


There are some funky things in this dataset that we will ignore for our analysis. This includes the first and second ID and diagnosis columns and the "X" column.
```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc.df[,3:32])
head(wisc.data)
```

Q1. How many patients do we have data for? 569
```{r}
nrow(wisc.data)
```

Q2. How many of the observations have a malignant diagnosis? B = 357 & M = 212
```{r}
table(wisc.df$diagnosis)
```

Q3. How many variables/features in the data are suffixed with "_mean"?
For this we will use `grep()`
```{r}
#colnames(wisc.data) -> have to specify we're looking for "_mean" in the column names
grep("_mean", colnames(wisc.data))
```
```{r}
# To determine how many columns have "_mean" 
length(grep("_mean", colnames(wisc.data)))
```

##Principal Component Analysis

Before we turn to PCA we need to think, or consider, whether we should scale our input. 

- The input variables use different units of measurement.
- The input variables have significantly different variances.
Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the `colMeans()` and `apply()` functions like you’ve done before.
```{r}
# Check column means and standard deviations
#colMeans(wisc.data)
# 1 for rows or 2 for columns -> calculate the SD for every column
round( apply(wisc.data,2,sd ), 2)
```


Looks like we need to set scale = TRUE 
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale = TRUE )
# Look at summary of results
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
We can look at the summary above. The 1st PC captures 44.27%  of the original variance.72.6% is captured in the first 3 PCs.

Let's make some figures.
```{r}
biplot(wisc.pr)
```

We need to do our own PC1 vs PC2 plot, and let's color by the diagnosis.

```{r}
attributes(wisc.pr)
```

```{r}
# Scatter plot observations by components 1 and 2

plot( wisc.pr$x[,1] , wisc.pr$x[,2], col = wisc.df$diagnosis , 
     xlab = "PC1", ylab = "PC2")
```

###Cluster in PC space
First let's see if we can cluster the original data.
```{r}
kmeans(wisc.data, centers = 2, nstart = 20)
```

```{r}
wisc.hc <- hclust(dist(wisc.data))
plot(wisc.hc)

```

This does not look good, so we'll try and combine the results of PCA with clustering.

Let’s see if PCA improves or degrades the performance of hierarchical clustering.

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.
```{r}
# Clustering PCA1 to PCA3
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:3]), method="ward.D2")
plot(wisc.pr.hclust)
```

To get our clusters out of this tree we need to cut it with the `cutree()` function.
```{r}
grps3 <- cutree(wisc.pr.hclust, k=2)
table(grps3)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=grps3)
```

We can use the `table()` function to compare the $diagnosis vector with our cluster results vector
```{r}
# This shows how many from one category cross over to another category
table(grps3, wisc.df$diagnosis)
```

##Prediction

We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.
```{r}
new <- read.csv("new_samples.csv")
new
```

Use the `predict()` function with or previous PCA model and new data.
```{r}
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = wisc.df$diagnosis)
points(npc[,1], npc[,2], col= "blue", pch = 15, cex = 3)
text(npc[,1], npc[,2], labels = c(1,2), col = "white")
```










